Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-17 manuscript.tex
first stab at amendments

Solutions to the linearised Maxwell-Vlasov system of equations are found in
parallel and perpendicular wavenumber space, $(k_\parallel, k_\bot)$, for a
ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions. Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region, see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV. The complex frequency
solutions, $\omega$, along the fast Alfv{\'e}n branch are found on a grid of $(k_\parallel, k_\bot)$.

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 in the same reference. The electrons and deuterons have
temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds a major radius of $4$ m.
The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
This distribution function is parameterised by the drift
velocity parallel to the magnetic field $u_\parallel$, the drift speed 
perpendicular to the magnetic
field $u_{\perp}$, and the thermal velocity in both directions $v_{th}$. The
thermal spread of the alpha particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ given the distribution functions of all the species present.
The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$, Eq. \ref{eq:dielectrictensor} is calculated for each species.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch} except there is an additional black trace in the
upper panel, which represents the measured JET ICE intensity signal; the continuous
trace from Fig. 2 in Ref. \cite{Cottrell1993}.
The observations of ICE intensity are plotted on a log scale scaled to match
the range of the linear MCI growth rates.
Direct comparison of log-scale
ICE observations and linear growth rates is justified to the extent that
nonlinear PIC simulations indicate that the linearly unstable modes rising
as $\exp(\gamma t)$, where $\gamma$ is a mode's growth rate,
cease to grow once the fastest growing mode becomes
nonlinear\cite{Cook2013}.
Therefore under the assumption that all modes start growing in unison from
identical seed amplitudes then the log of a mode's amplitude is
proportional to its growth rate.
There are several points worth noting about this figure. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148\cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step change in signal strength from $\omega < 6$ to $\omega > 6$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\parallel,
k_\perp)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron
frequency, which makes it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers. Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.

>>>>

Solutions to the linearised Maxwell-Vlasov system of equations are found in
parallel and perpendicular wavenumber space, $(k_\parallel, k_\bot)$, for a
ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions.

The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised  by the drift
velocity parallel to the magnetic field $u_\parallel$, the drift speed 
perpendicular to the magnetic
field $u_{\perp}$, and the thermal velocity in both directions $v_{th}$. The
thermal spread of the alpha particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET, it is worth noting that they are both of
order $\rho^*$ and have been neglected in these calculations where $u_\bot$ is
always large.

Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) u_0 \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region, see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV.

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 in the same reference. The electrons and deuterons have
temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds a major radius of $4$ m.

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ given the distribution functions of all the species present
along the fast Alfv{\'e}n branch  on a grid of $(k_\parallel, k_\bot)$.
The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$, Eq. \ref{eq:dielectrictensor} is calculated for each species.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch} except there is an additional black trace in the
upper panel, which represents the measured JET ICE intensity signal; the continuous
trace from Fig. 2 in Ref. \cite{Cottrell1993}.
The observations of ICE intensity are plotted on a log scale scaled to match
the range of the linear MCI growth rates.
Direct comparison of log-scale
ICE observations and linear growth rates is justified to the extent that
nonlinear PIC simulations indicate that the linearly unstable modes rising
as $\exp(\gamma t)$, where $\gamma$ is a mode's growth rate,
cease to grow once the fastest growing mode becomes
nonlinear\cite{Cook2013}.
Therefore under the assumption that all modes start growing in unison from
identical seed amplitudes then the log of a mode's amplitude is
proportional to its growth rate.

Although this simplication neglects possibly important effects, such as particle sink
and source terms, and their influence on the nonlinear amplitudes of the
observed ICE signals it is still possible to obtain important information
regarding the signals' structure and overall trends.

There are several points worth noting about this figure. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Ref. \cite{Dendy1994} (also
Fig. 9 Ref. \cite{Dendy1995}) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148\cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step change in signal strength from $\omega < 6\Omega_i$ to $\omega > 6\Omega_i$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\parallel,
k_\perp)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron
frequency, which makes it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers. Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.
